---
output: 
  html_document:
    keep_md: true
---

This is the code for Project of week 2 of Reproducible Research by Coursera

We will now load the data here which is in csv format

# 1. Start of Loading the data and pre process of the data
```{r, echo=TRUE}
library("data.table")
library(ggplot2)

activity <- read.csv("activity.csv")
head(activity)
summary(activity)
```
# # End of Loading the data and pre process of the data


# 2. Start of Mean Total Number of Steps taken per day

## For that we need couple of steps before we calculate the total number of steps.

## The first is to change the date type to class, then NA to logical vector and create a clean base for later mean caclulations

```{r, echo=TRUE}
# change date type to date class 
activity$date <- as.Date(as.character(activity$date))
# create NA logical vector
activityNA <- is.na(activity$steps)
# create clean base for later mean calculations
clean_activity <- activity[!activityNA,]
```

# 2a. What is mean total number steps taken per day?

## Now we will calculate the totl number of steps by using the aggregeate function 

```{r, echo=TRUE}
# aggregate the daily sum of steps taken
DataByDay <- aggregate(activity$steps, by=list(activity$date), sum)
# adjust column names
names(DataByDay)[1] ="date"
names(DataByDay)[2] ="totalsteps"
# top 10 of Summed Steps by day
head(DataByDay,10)
```

## 2b. Now we are going to plot the histogram of number of steps taken using ggplot. And This will be the question number 2 of the project.

```{r, echo=TRUE}
ggplot(DataByDay, aes(x = totalsteps)) +
  geom_histogram(fill = "steelblue", binwidth=1000) +
  labs(title = "Total Daily Steps", x = "Steps", y = "Frequency")
dev.print(png, file = "plot1.png", width = 480, height = 480)
png(file = "plot1.png", bg = "transparent")
dev.off()
```

## 2c. Calculate and report the mean and median of the total number of steps taken per day

```{r, echo=TRUE}
# Mean of steps taken per day
mean(DataByDay$totalsteps,na.rm=TRUE)
#median of steps taken per day
median(DataByDay$totalsteps,na.rm=TRUE)
```


# 3. What is the average daily activity pattern

## 3a. Making Time Series plot of 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r, echo=TRUE}
No_NA <- activity[!activityNA,]
mean_data_Interval <- aggregate(No_NA$steps, by=list(No_NA$interval), mean)
# set the column names
names(mean_data_Interval)[1] ="interval"
names(mean_data_Interval)[2] ="steps"

ggplot(mean_data_Interval, aes(x = interval, y=steps)) +
  labs(title = "Sum of Steps by Interval", x = "interval", y = "steps")+
  geom_line(color="red")
dev.print(png, file = "plot2.png", width = 480, height = 480)
png(file = "plot2.png", bg = "transparent")
dev.off()
```

## 3b. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r, echo=TRUE}
maxsteps <- mean_data_Interval[which.max(mean_data_Interval$steps),]
maxsteps
```


# 4. Imputing the Missing Values

## 4a, 4b and 4c. Calculating and reporting the total number of missing values in the dataset (i.e. the total number of rows with NAs
```{r, echo=TRUE}
missing_values <- sum(activityNA)
missing_values


## Devise a strategy for filling in all of the missing values in the dataset. 

### New Data Set
activity2 <- activity
### Removing NA values from new dataset
No_NA_activity <- activity2[is.na(activity2$steps),]
clean_activity <- activity2[!is.na(activity2$steps),]


# calculating the mean of new dataset
Mean_activity_new <- aggregate(clean_activity$steps, by=list(clean_activity$interval), sum)
names(Mean_activity_new)[1] ="interval"
names(Mean_activity_new)[2] ="steps"


# Imputing
activity2 <- activity
missing_values <- is.na(activity2$steps)
meanVals <- tapply(clean_activity$steps, clean_activity$interval, mean, na.rm=TRUE, simplify=TRUE)
activity2$steps[missing_values] <- meanVals[as.character(activity2$interval[missing_values])]
# count of NA values
sum(is.na(activity2$steps))


```

## 4d. Making the histogram of total number os steps tke and calculating the mean and median of total number of steps taken per day.

```{r, echo=TRUE}
final_sum_data_by_data <- aggregate(activity2$steps, by=list(activity2$date), sum)

names(final_sum_data_by_data)[1] ="date"
names(final_sum_data_by_data)[2] ="totalsteps"
head(final_sum_data_by_data,10)

ggplot(final_sum_data_by_data, aes(x = totalsteps)) +
  geom_histogram(fill = "steelblue", binwidth=1000) +
  labs(title = "Total Daily Steps", x = "Steps", y = "Frequency")
dev.print(png, file = "plot3.png", width = 480, height = 480)
png(file = "plot3.png", bg = "transparent")
dev.off()

mean(final_sum_data_by_data$totalsteps)
median(final_sum_data_by_data$totalsteps)

```

### The mean remains same but median value increased 1.19 steps

### The over all effect is that data is pushed towards mean

# 5. Are there differences in activity patterns between weekdays and weekends?

## 5a. Creating new factor variable in the dataset with two levels "weekday" and "weekend"

```{r, echo=TRUE}
activity2$weekday <- weekdays(activity2$date)
activity2$weekend <- ifelse (activity2$weekday == "Saturday" | activity2$weekday == "Sunday", "Weekend", "Weekday")
head(activity2,5)
```

## 5b. Making a panel plot containing the time series plot of 5 minute interval 

```{r, echo=TRUE}
combined_weekday_weekend_mean <- aggregate(activity2$steps, by=list(activity2$weekend, activity2$interval), mean)
names(combined_weekday_weekend_mean)[1] ="weekend"
names(combined_weekday_weekend_mean)[2] ="interval"
names(combined_weekday_weekend_mean)[3] ="steps"

ggplot(combined_weekday_weekend_mean, aes(x = interval, y=steps, color=weekend)) +
  geom_line() +
  facet_grid(weekend ~ .) +
  labs(title = "Mean of Steps by Interval", x = "interval", y = "steps")
dev.print(png, file = "plot4.png", width = 480, height = 480)
png(file = "plot4.png", bg = "transparent")
dev.off()
```







